# Phylogenetic Generalized Linear Models

Load the same libraries and datasets as in the prior chapter.

```{r echo=FALSE, eval=FALSE}
#Libraries required
library(ape) #contains read.nexus. plot.phylo
library(geiger)#fitContinuous
library(phytools) #vcvPhylo
library(nlme)
library(phylolm)
```

## Phylogenetic Linear Models

In our phylogenetic linear model we create an important macroevolution hypothesis: Flower length is determined by the types of pollinators that visit the flower. 

EXTREMELY IMPORTANT HERE = The main assumption of these models is that the ERRORS (not the response variable) have a normal distribution with a vector of zeros as the mean and variance-covariance matrix sigma^2 *C. Remember C is determined by the structure of the tree as we d calculated it in the continuous trait (Brownian motion) tutorial.

### Testing the hypothesis

How would you frame then a biological hypothesis from these observations?

Before typing a model it would be great to think about what we are trying to achieve.

Now use the following code and fill in the necessary arguments.

```{r eval=FALSE}
#not log.flowerlength?
plm.flowerlength <- phylolm(formula, data=, phy=, model="BM")
```

Interpret this model, what is happening? what is significant and how
```
summary(plm.flowerlength)
```

What is happening with the residuals of this linear model

```
hist(plm.flowerlength$res)
qqnorm(plm.flowerlength$res)
```

Now what happens if the predictor is log(Length)?

```{r eval=FALSE}
plm.flowerlength2 <- phylolm(formula, data=, phy=, model="BM")
```

Using the same commands as above,
interpret this model, what is happening? what is significant and how

## What happens if the model of evolution is not Brownian motion but something else?

```{r echo=TRUE, eval=FALSE}
plm.flowerlength3<-phylolm(Length~ P1+P2+P3+ P4, data=pole_dataset, phy=pole_tree, model="OUrandomRoot")
```

What is this model?

```
summary(plm.flowerlength3)
```

interpret this model, what is happening? what is significant and how

```
plot(plm.flowerlength3)
hist(plm.flowerlength3$res)
qqnorm(plm.flowerlength3$res)
```

Overall which is the best model and why?  Argue statistically but also think about flower length.

