# Phylogenetic Generalized Linear Models

Packages used for this section:
---
```{r echo=TRUE, eval=FALSE}
#Libraries required
library(ape) #contains read.nexus. plot.phylo
library(geiger)#fitContinuous
library(phytools) #vcvPhylo
library(nlme)
library(phylolm)
```

## Reading dataset and phylogenetic tree

Data is a simplify version of the data and models we published here
Landis, J.B., Bell, C.D., Hernandez, M., Zenil-Ferguson, R., McCarthy, E.W., Soltis, D.E. and Soltis, P.S., 2018. Evolution of floral traits and impact of reproductive mode on diversification in the phlox family (Polemoniaceae).Molecular phylogenetics and evolution, 127, pp.878-890.
We have a tree with 165 tips from the Polimoniaceae family (Phlox). They grow in California.  We are interested in the rate of evolution of flower length

```{r echo=TRUE, eval=FALSE}
## Read the data
setwd("~/Dropbox/AndeanWorkshop")
pole_tree<-read.nexus("pole_tree.nex")
```

This is a dataset of traits and pollinators visiting the flowers on the tips of the tree

```{r echo=TRUE, eval=FALSE}
pole_dataset<-read.csv("pole_dataset.csv",header=TRUE)
head(pole_dataset)
rownames(pole_dataset) <- pole_tree$tip.label 
head(pole_dataset)
## Description
## Length- flower length (Response)
## P1- butterflies visit or not (Predictor)
## P2- hawkmoths visit or not (Predictor)
## P3- bee flies visit or not (Predictor)
## P4- bees visit or not (Predictor)
```


##Phylogenetic Linear Models

Jacob Landis measured the flower width and length and observed which pollinators come visit the flowers. Jacob was interested in understanding if we can predict flower shape based on pollinator type. For example, flowers that hummingbirds visit might be longer and narrower than flowers that are visited by bees.

I wanted to create a model where the response is the flower shape and the predictor is what type of pollinator visits the flower

```{r echo=TRUE, eval=FALSE}
lengthplot <- setNames(pole_dataset$Length,rownames(pole_dataset)) #Stop to see what this line of code is doing

# Plotting by hummingbird pollination
mycolors <- rep("blue",length(pole_dataset$Length))
mycolors[which(pole_dataset$P1==1)]="red"

plotTree.barplot(tree=pole_tree,x=lengthplot,args.plotTree=list(ftype="off"),
    args.barplot=list(col=mycolors,border=mycolors,xlab=""))
legend(x=3, y=100,legend=c("Hummingbird pollinated","Other pollination"),pch=22,cex=0.5,pt.bg=c("red","blue"),
    box.col="transparent")

```


In our  phylogenetic linear model we make create an important macroevolution hypothesis: Flower length is determined by the types of pollinators that visit the flower. 

EXTREMELY IMPORTANT HERE = The main assumption of these models is that the ERRORS (not the response variable) have a normal distribution with a vector of zeros as the mean and variance-covariance matrix sigma^2 *C. Remember C is determined by the structure of the tree as we d calculated it in the continuous trait (Brownian motion) tutorial.

### Testing the hypothesis

How would you frame then a biological hypothesis from these observations?

Before typing a model it would be great to think about what we are trying to achieve.

```{r eval=FALSE}
#not log.flowerlength?
plm.flowerlength<-phylolm(formula, data=, phy=, model="BM")
## interpret this model, what is happening? what is significant and how
summary(plm.flowerlength)

# What is happening with the residuals of this linear model
hist(plm.flowerlength$res)
qqnorm(plm.flowerlength$res)
```


Now what happens if the predictor is log(Length)?
```{r eval=FALSE}
#Complete the model
plm.flowerlength2<-phylolm(formula, data=, phy=, model="BM")

summary(plm.flowerlength2)
## interpret this model, what is happening? what is significant and how

hist(plm.flowerlength2$res)
qqnorm(plm.flowerlength2$res)
```



# What happens if the model of evolution is not Brownian motion but something else?

```{r echo=TRUE, eval=FALSE}
plm.flowerlength3<-phylolm(Length~ P1+P2+P3+ P4, data=pole_dataset, phy=pole_tree, model="OUrandomRoot")
#What is this model?
 summary(plm.flowerlength3)
## interpret this model, what is happening? what is significant and how

plot(plm.flowerlength3)
hist(plm.flowerlength3$res)
qqnorm(plm.flowerlength3$res)
```

Overall which is the best model and why?  Argue statistically but also think about flower length.

