#Discrete variable modeling

In this section, we will work on modeling a continuous trait and understanding what Brownian motion is.
```{r echo=TRUE, eval=FALSE}
#Libraries required
library(ape) #contains read.nexus. plot.phylo
library(geiger)#fitContinuous
library(phytools) #vcvPhylo
library(nlme)
library(phylolm)
library(expm)
library(corHMM)
```

## Reading dataset and phylogenetic tree
This data comes from Landis et al.2018. This is a tree with 165 tips from the Polimoniaceae family (Phlox). They grow in California.  We are interested in the rate of evolution of flower length

```{r echo=TRUE, eval=FALSE}
## Read the data
## Read the data
setwd("~/Desktop")
pole_tree<-read.nexus("pole_tree.nex")
pole_tree
```

This is a dataset of traits and pollinators visiting the flowers on the tips of the tree

```{r echo=TRUE, eval=FALSE}
pole_dataset<-read.csv("pole_dataset.csv",header=TRUE)
head(pole_dataset)
rownames(pole_dataset) <- pole_tree$tip.label 
head(pole_dataset)
```


### Plotting discrete data
There are multiple ways to plot discrete data on a phylogeny. Multiple packages too (``ape``,``ggtree``, but we will use ``phytools``).


Using ``phytools`` we will plot a tree with tip data as colors.
```{r echo=TRUE, eval=FALSE}
# Plotting by hummingbird pollination
plotTree(pole_tree,ftype="i",offset=0.6,fsize=0.3)
aux<-as.factor(pole_dataset$P1)
aux<-setNames(aux, pole_tree$tip.label)

#Need to create a matrix for the pie chart
discrete.matrix<-to.matrix(aux,levels(aux))

tiplabels(pie=discrete.matrix,piecol=c("blue","red"),cex=0.1)
legend("topleft",levels(aux),pch=21,pt.bg=c("blue","red"),
    pt.cex=2.2)
```

Notice that the function ``tiplabels()`` can draw pies, meaning it can draw more than one state for a tip.
## Discrete character evolution

For models of discrete character evolutin we will use a mathematical model called *Continuous-time Markov chains (CTMC)*. CTMCs are stochastic (random) processes that allow us to follow the evolution of a trait(s) of interest over the branches of the tree. In this example we will focus on the evolution of hummingbird pollination.

CTMCs are usually denoted in mathematical notation as $$\{X(t), t\geq 0 \}$$; the stochastic process $$X(t)$$ follows the pollination throughout time $$t$$. Time is continuous and it is measured using branch length of our phylogeny. For discrete traits the process $$X(t)$$ takes values  in the natural numbers $$0,1,2,...$$.  CTMCs have many (cool) mathematical properties, but one of the most important  property is the *Markovian property* that states that  the probability distribution of future states of the process conditional on both past and present states depends only upon the present state.


Mathematically, CTMC can is defined through a circle and arrow diagram and that gets converted into a matrix that we called the Q-matrix.


### What is a Q-matrix?
In mathematics, the Q-matrix is the derivative of the probability matrix ($$P'(t)=P(t)Q$$). The  elements outside of the diagonal of the $$Q$$ matrix are transition rates and in the diagonal we have the negative sum of all the elements of the row. The row adds to zero this is because probability matrices rows add to 1 so the derivative of the constant 1 is zero.

Let's stop here to think through this problem by drawing it. 


## Modeling using CTMCs
In phylogenetic comparative methods we have different types of CTMCs. Examples of these models are:

1. Evolution of tempo of a continuous trait (past lesson) -> BM, OU, EB, Levy, implemented in ``geiger``, ``phylolm``, ``bayou`` R packages.
2. Evolution of the tempo between states-> Markov models (Mkn) implemented in ``phytools``, ``corHMM``
3. Evolution of the tempo between states considering other potential sources promoting evolution-> Hidden state Markov models (HMM) implemented in ``corHMM``
4. Phylogeographic models that include dispersal or extirpation and cladogenesis -> DEC, DECj, DiVA implemented in ``Biogeobears``
5. Diversification (speciation and extinction) that changes over time -> Birth and death models implemented in ``castor``
6. Diversification (speciation and extinction) that changes based on clades or density -> Birth and death models implemented in ``BAMM``, ``CLADS``
7. Diversification (speciation and extinction) that changes based on state values-> State-dependent diversification models for example BiSSE, HiSSE implemented in ``hisse``
8. Diversification (speciation and extinction) that changes based on biogeography or cladogenesis -> State-dependent diversification models for example GeoSSE implemented in ``hisse``

Learning about CTMCs is then critical for us to develop hypotheses about trait evolution, diversification..

### Modeling the transition rates between two states

We are going to use a CTMC model called Mk2 (Markov model with two states). Let's take a minute to draw define the model. We are interested in the rate of evolution of hummingbird pollination in Phlox species.

```{r echo=TRUE, eval=FALSE}
data.discrete<-cbind(pole_tree$tip.label, pole_dataset$P1)
mk2<-corHMM(phy=pole_tree, data=data.discrete, rate.cat=1)
mk2 #Let's read and interpret the results carefully
plotMKmodel(mk2)
```

Assumption: Equal rates

```{r echo=TRUE, eval=FALSE}
LegendAndRateMat<-getStateMat4Dat(data.discrete)
RateMat<-LegendAndRateMat$rate.mat
pars2equal<-list(c(1,2))
constrained.mat<-equateStateMatPars(RateMat,pars2equal)
constrained.mat
mk2_equalrates<-corHMM(phy=pole_tree, data=data.discrete, rate.cat=1, rate.mat=constrained.mat)
mk2_equalrates
plotMKmodel(mk2_equalrates)
```

### Model comparison 

It is important to use model comparison in this context to investigate whether the rates of evolution between the two states are equal or different. In likelihood context we use the Akaike Information Criterion known as AIC that is defined as
$$AIC= -2 \log(\textrm{Likelihood evaluated in the MLE)}+ 2\times \textrm{number of parameters}$$
AIC allows us to compare between models with different number of parameters because it corrects for it. We want to choose for the model with the smallest AIC. A difference of 2 or larger between two AIC values is considered as evidence in favor of a given model.


So knowing this, which model of evolution would you choose for this example?

### Stochastic mapping
Visualizing the history of evolution of a trait using stochastic maps. 

```{r echo=TRUE, eval=FALSE}
phy = mk2_equalrates$phy
data = mk2_equalrates$data
model = mk2_equalrates$solution
model[is.na(model)] <- 0
diag(model) <- -rowSums(model)
# run get simmap (can be plotted using phytools)
simmap <- makeSimmap(tree=phy, data=data, model=model, rate.cat=1, nSim=10, nCores=1)

# we import phytools plotSimmap for plotting
dev.off()
plotSimmap(simmap[[1]], fsize=.4)
colors<-setNames(c("black","red"),0:1)
dev.off()
densityTree(simmap,method="plotSimmap",lwd=8,nodes="intermediate", ylim=c(1,170),colors=colors, compute.consensus=FALSE, fsize=0.2, show.axis=FALSE, fix.depth=TRUE)
```
,
### Modeling hidden states

Often, the transitions between to states are associated with other changes in the history of the trait itself. Some of those changes we can measure i.e., polyploidy vs. breeding systems, but some others are hard to quantify or hard to measure. Because many things could have happen that are linked to transitions between states we need to model using what we call hidden states. 

The hidden states model is a mathematical tool that allow us to parse the main effect of the trait of interest, by removing other potential sources of evolution in the transitons.


We will stop here and draw what these hidden states are!


How do we implement them in R?
Two different assumptions
```{r echo=TRUE, eval=FALSE}
hmm2_2_equal<- corHMM(phy = pole_tree, data = data.discrete, rate.cat = 2, model = "SYM", get.tip.states = TRUE)
hmm2_2_equal#Let's read and interpret the results carefully
plotMKmodel(hmm2_2_equal)

hmm2_2_different<- corHMM(phy = pole_tree, data = data.discrete, rate.cat = 2, model = "ARD", get.tip.states = TRUE)
hmm2_2_different#Let's read and interpret the results carefully
plotMKmodel(hmm2_2_different)

```

What is a better? What is the overall best model??
